---
title: "Predicted untranslated regions"
author: "Lindsay Clark"
format: html
editor: visual
---

## Overview

Since not all genes have UTRs defined in the genome annotation, I will define predicted UTRs in which to search.

## Import

```{r}
library(GenomicRanges)
library(dplyr)
```

```{r}
ortho_df <- read.csv("../resources/vivax_yoelii_orthologs_2026-02-11.csv")

gff_pv <- rtracklayer::import("../resources/PlasmoDB-68_PvivaxP01.gff.gz")
gff_py <- rtracklayer::import("../resources/PlasmoDB-68_Pyoeliiyoelii17X.gff.gz")
```

## P. vivax

Import UTRs, document their width

```{r}
gff_pv_cds <- gff_pv[gff_pv$type == "CDS"]
gff_pv_mrna <- gff_pv[gff_pv$type == "mRNA"]
utr_df <- data.frame(Transcript = gff_pv_mrna$ID)
utr_df$Gene <- sub("\\.[[:digit:]]+$", "", utr_df$Transcript)

gff_pv_5utr <- gff_pv[gff_pv$type == "five_prime_UTR"]
gff_pv_3utr <- gff_pv[gff_pv$type == "three_prime_UTR"]

temp_plasmo <- data.frame(Transcript = unlist(gff_pv_5utr$Parent),
                          Width_5UTR_PlasmoDB = width(gff_pv_5utr))
utr_df <- left_join(utr_df, temp_plasmo, by = "Transcript")
utr_df$Width_5UTR_PlasmoDB[is.na(utr_df$Width_5UTR_PlasmoDB)] <- 0L

temp_plasmo <- data.frame(Transcript = unlist(gff_pv_3utr$Parent),
                          Width_3UTR_PlasmoDB = width(gff_pv_3utr))
utr_df <- left_join(utr_df, temp_plasmo, by = "Transcript")
utr_df$Width_3UTR_PlasmoDB[is.na(utr_df$Width_3UTR_PlasmoDB)] <- 0L

rm(temp_siegel, temp_plasmo)

gff_pv_mrna <- gff_pv[gff_pv$type == "mRNA"]
mtch <- match(utr_df$Transcript, gff_pv_mrna$ID)
utr_df$seqnames <- as.character(seqnames(gff_pv_mrna)[mtch])
utr_df$strand <- as.character(strand(gff_pv_mrna)[mtch])

gff_pv_cds <- gff_pv[gff_pv$type == "CDS"]
temp <- data.frame(Gene = unique(gff_pv_cds$gene_id))
temp$cds_start <- sapply(temp$Gene,
                         \(x){
                             min(start(gff_pv_cds[gff_pv_cds$gene_id == x]))
                         })
temp$cds_end <- sapply(temp$Gene,
                       \(x){
                           max(end(gff_pv_cds[gff_pv_cds$gene_id == x]))
                       })
utr_df <- left_join(utr_df, temp, by = "Gene")
```

Get the median UTR widths for known UTRs. This is 808 nt for 5' UTRs and 688 for 3' UTRs.

```{r}
widths_5utr <- tapply(utr_df$Width_5UTR_PlasmoDB, utr_df$Transcript, sum)
med5 <- median(widths_5utr[widths_5utr > 0])

widths_3utr <- tapply(utr_df$Width_3UTR_PlasmoDB, utr_df$Transcript, sum)
med3 <- median(widths_3utr[widths_3utr > 0])
```

Get predicted coordinates

```{r}
utr_df_sub5 <- utr_df[utr_df$Width_5UTR_PlasmoDB == 0,]
utr_df_sub3 <- utr_df[utr_df$Width_3UTR_PlasmoDB == 0,]

temp <- utr_df_sub5$strand == "+"
utr_df_sub5$utr5_start[temp] <- utr_df_sub5$cds_start[temp] - med5
utr_df_sub5$utr5_end[temp] <- utr_df_sub5$cds_start[temp] - 1L

temp <- utr_df_sub5$strand == "-"
utr_df_sub5$utr5_end[temp] <- utr_df_sub5$cds_end[temp] + med5
utr_df_sub5$utr5_start[temp] <- utr_df_sub5$cds_end[temp] + 1L

temp <- utr_df_sub3$strand == "+"
utr_df_sub3$utr3_end[temp] <- utr_df_sub3$cds_end[temp] + med3
utr_df_sub3$utr3_start[temp] <- utr_df_sub3$cds_end[temp] + 1L

temp <- utr_df_sub3$strand == "-"
utr_df_sub3$utr3_start[temp] <- utr_df_sub3$cds_start[temp] - med3
utr_df_sub3$utr3_end[temp] <- utr_df_sub3$cds_start[temp] - 1L

utr_df_sub5$utr5_start[utr_df_sub5$utr5_start < 0] <- 0L
utr_df_sub3$utr3_start[utr_df_sub3$utr3_start < 0] <- 0L
```

Build GRanges objects for these

```{r}
pred_pv_5utr <- GRanges(seqnames = utr_df_sub5$seqnames,
                        ranges = IRanges(utr_df_sub5$utr5_start,
                                         utr_df_sub5$utr5_end),
                        strand = utr_df_sub5$strand)
pred_pv_5utr$gene_id <- utr_df_sub5$Gene
pred_pv_5utr$type <- "five_prime_UTR"

pred_pv_3utr <- GRanges(seqnames = utr_df_sub3$seqnames,
                        ranges = IRanges(utr_df_sub3$utr3_start,
                                         utr_df_sub3$utr3_end),
                        strand = utr_df_sub3$strand)
pred_pv_3utr$gene_id <- utr_df_sub3$Gene
pred_pv_3utr$type <- "three_prime_UTR"
```

## P. yoelii

In the P. yoelii GFF we have exons and CDS listed, but not UTRs. I tried to define them by subtraction but there's really not much there. Just predict UTRs based on the median sized from P. vivax.

```{r}
gff_py_cds <- gff_py[gff_py$type == "CDS"]
gff_py_mrna <- gff_py[gff_py$type == "mRNA"]
utr_df_py <- data.frame(Transcript = gff_py_mrna$ID)
utr_df_py$Gene <- sub("\\.[[:digit:]]+$", "", utr_df_py$Transcript)
utr_df_py$seqnames <- as.character(seqnames(gff_py_mrna))
utr_df_py$strand <- as.character(strand(gff_py_mrna))
utr_df_py$cds_start <- NA_integer_
utr_df_py$cds_end <- NA_integer_
utr_df_py$utr5_start <- NA_integer_
utr_df_py$utr5_end <- NA_integer_
utr_df_py$utr3_start <- NA_integer_
utr_df_py$utr3_end <- NA_integer_

temp_parent <- as.character(unlist(gff_py_cds$Parent))

for(i in seq_len(nrow(utr_df_py))){
    theserow <- which(temp_parent == utr_df_py$Transcript[i])
    utr_df_py$cds_start[i] <- min(start(gff_py_cds[theserow]))
    utr_df_py$cds_end[i] <- max(end(gff_py_cds[theserow]))
}

topstrd <- utr_df_py$strand == "+"
utr_df_py$utr5_start[topstrd] <- utr_df_py$cds_start[topstrd] - med5
utr_df_py$utr5_end[topstrd] <- utr_df_py$cds_start[topstrd] - 1L

utr_df_py$utr5_end[!topstrd] <- utr_df_py$cds_end[!topstrd] + med5
utr_df_py$utr5_start[!topstrd] <- utr_df_py$cds_end[!topstrd] + 1L

utr_df_py$utr3_end[topstrd] <- utr_df_py$cds_end[topstrd] + med3
utr_df_py$utr3_start[topstrd] <- utr_df_py$cds_end[topstrd] + 1L

utr_df_py$utr3_start[!topstrd] <- utr_df_py$cds_start[!topstrd] - med3
utr_df_py$utr3_end[!topstrd] <- utr_df_py$cds_start[!topstrd] - 1L
```

Generate GRanges objects of predicted UTRs

```{r}
pred_py_5utr <- GRanges(seqnames = utr_df_py$seqnames,
                        ranges = IRanges(utr_df_py$utr5_start,
                                         utr_df_py$utr5_end),
                        strand = utr_df_py$strand)
pred_py_5utr$gene_id <- utr_df_py$Gene
pred_py_5utr$Parent <- utr_df_py$Transcript
pred_py_5utr$type <- "five_prime_UTR"

pred_py_3utr <- GRanges(seqnames = utr_df_py$seqnames,
                        ranges = IRanges(utr_df_py$utr3_start,
                                         utr_df_py$utr3_end),
                        strand = utr_df_py$strand)
pred_py_3utr$gene_id <- utr_df_py$Gene
pred_py_3utr$Parent <- utr_df_py$Transcript
pred_py_3utr$type <- "three_prime_UTR"
```

Save predicted UTRs (not including anything that is annotated)

```{r}
save(pred_pv_5utr, pred_pv_3utr, pred_py_5utr, pred_py_3utr,
     file = "../results/R_objects/pred_UTR_GRanges_2026-03-25.RData")
```
