---
title: "Motif enrichment statistical tests"
author: "Lindsay Clark"
format: html
editor: visual
---

## Overview

In this notebook we test for the presence of motifs within transcripts, and perform downstream analysis.

```{r}
library(Biostrings)
library(GenomicRanges)
library(dplyr)
library(ggplot2)
library(ggrepel)
library(fgsea)
library(edgeR)
library(ggtext)
library(jsonlite)
library(plotrix)
library(showtext)
library(universalmotif)

font_add("Arial", regular = "arial.ttf", bold = "arialbd.ttf",
         italic = "ariali.ttf", bolditalic = "arialbi.ttf")
showtext_auto()
```

```{r}
ortho_df <- readRDS("../results/R_objects/expressed_orthologs.rds")

gff_pv <- rtracklayer::import("../resources/PlasmoDB-68_PvivaxP01.gff.gz")
gff_py <- rtracklayer::import("../resources/PlasmoDB-68_Pyoeliiyoelii17X.gff.gz")

load("../results/R_objects/pred_UTR_GRanges_2026-03-25.RData")

fasta_pv <- readDNAStringSet("../resources/PlasmoDB-68_PvivaxP01_Genome.fasta")
fasta_py <- readDNAStringSet("../resources/PlasmoDB-68_Pyoeliiyoelii17X_Genome.fasta")

names(fasta_pv) <- sub(" .*", "", names(fasta_pv))
names(fasta_py) <- sub(" .*", "", names(fasta_py))
```

## Frequency of motifs in 3' UTRs compared to random chance

Extract sequences of 3' UTRs in both species. We can use `BSgenome::getSeq` to properly flip sequences on the minus strand.

```{r}
gff_pv_3utr <- gff_pv[gff_pv$type == "three_prime_UTR"]

fix_widths <- function(gr, fasta){
    wd <- width(fasta)
    names(wd) <- names(fasta)
    max_end <- wd[as.character(seqnames(gr))]
    needs_fix <- end(gr) > max_end
    end(gr)[needs_fix] <- max_end[needs_fix]
    start(gr)[start(gr) < 1L] <- 1L
    return(gr)
}

pred_pv_3utr <- fix_widths(pred_pv_3utr, fasta_pv)

seq_pv_3utr_1 <- BSgenome::getSeq(fasta_pv, gff_pv_3utr)
names(seq_pv_3utr_1) <- sub("\\..*", "", sub("^utr_", "", gff_pv_3utr$ID))
seq_pv_3utr_2 <- BSgenome::getSeq(fasta_pv, pred_pv_3utr)
names(seq_pv_3utr_2) <- pred_pv_3utr$gene_id

seq_pv_3utr <- c(seq_pv_3utr_1, seq_pv_3utr_2)
rm(seq_pv_3utr_1, seq_pv_3utr_2)

pred_py_3utr <- fix_widths(pred_py_3utr, fasta_py)

seq_py_3utr <- BSgenome::getSeq(fasta_py, pred_py_3utr)
names(seq_py_3utr) <- pred_py_3utr$gene_id
```

Eliminate stretches of N's in the vivax 3' UTRs

```{r}
seq_pv_3utr_split <- unlist(strsplit(seq_pv_3utr, "N"))
```

Function to run test for enrichment. I had previously used mononucleotide, then dinucleotide frequencies, across all provided sequences to predict the null rate for the motif. At a reviewer's recommendation I tried shuffling the sequences and then running the test, which I still found to be significant although not as much. This is probably because mononucleotide and dinucleotide frequencies are not uniform across the sequences. So, instead I am now using shuffled sequences to create my null expectation for the motif rate.

```{r}
test_motif_enrichment <- function(motif, seqs, shuffle_seeds = 1:10){
    motif <- gsub("U", "T", motif)
    
    # Get the number of sites big enough to have the motif
    temp <- width(seqs) - nchar(motif) + 1L
    temp[temp < 0L] <- 0L
    n_sites <- sum(temp)
    # Get the number of motif occurrences
    n_match <- sum(vcountPattern(motif, seqs, fixed = FALSE))
    
    # Shuffle the sequences
    seq_shuffled <- lapply(
        shuffle_seeds,
        \(x) shuffle_sequences(seqs, k = 2, nthreads = 4, rng.seed = x,
                               method = "euler")
        )
    n_match_shuffled <- sapply(
        seq_shuffled,
        \(x) sum(vcountPattern(motif, x, fixed = FALSE))
    ) |> sum()
    
    n_sites_shuffled <- n_sites * length(shuffle_seeds)
    
    # run Poisson test
    pt <- poisson.test(x = c(n_match, n_match_shuffled),
                       T = c(n_sites, n_sites_shuffled),
                       alternative = "greater")
    
    return(
        c(
            matches = n_match,
            sites = n_sites,
            shuffled_matches = n_match_shuffled,
            sites_shuffled = n_sites_shuffled,
            rate_ratio = unname(pt$estimate),
            p_value = pt$p.value
        )
    )
}
```

Test how the observed rate compares to the rate with shuffled sequences. One-sided since we only care about motifs more common than expected

- For UGUANNNUA in P. vivax, P = 9.6e-26. The rate ratio is 1.26.

- For UGUANNNUA in P. yoelii, P = 1; the rate ratio is near 1.

- For UGUAUAAUA in P. vivax, P = 1e-3, and the rate ratio is 1.33.

- For UGUAUAAUA in P. yoelii, P = 8.3e-7 and the rate ratio is 1.32.

```{r}
test_motif_enrichment("UGUANNNUA", seq_pv_3utr_split)

test_motif_enrichment("UGUANNNUA", seq_py_3utr)

test_motif_enrichment("UGUAUAAUA", seq_pv_3utr_split)

test_motif_enrichment("UGUAUAAUA", seq_py_3utr)
```

Test all UGUANNNUA and UGUANNUA motifs.

```{r}
letters <- c("A", "C", "G", "U")
all_motifs <- c(paste0(
    "UGUA",
    rep(letters, each = 16),
    rep(rep(letters, each = 4), times = 4),
    rep(letters, times = 16),
    "UA"
),
paste0(
    "UGUA",
    rep(letters, each = 4),
    rep(letters, times = 4),
    "UA"
)
)

all_tests_pv <- sapply(
    all_motifs,
    test_motif_enrichment,
    seqs = seq_pv_3utr_split
)

all_tests_py <- sapply(
    all_motifs,
    test_motif_enrichment,
    seqs = seq_py_3utr
)
```

Build data frame of results

```{r}
enrich_df <- data.frame(Motif = all_motifs,
                        Matches_pv = all_tests_pv["matches",],
                        Rate_ratio_pv = all_tests_pv["rate_ratio",],
                        Poisson_P_pv = all_tests_pv["p_value",],
                        Matches_py = all_tests_py["matches",],
                        Rate_ratio_py = all_tests_py["rate_ratio",],
                        Poisson_P_py = all_tests_py["p_value",])
```

See what Benjamini and Hochberg have to say about it

```{r}
padj <- p.adjust(c(enrich_df$Poisson_P_pv, enrich_df$Poisson_P_py),
                 method = "fdr")

enrich_df$FDR_global_pv <- padj[1:80]
enrich_df$FDR_global_py <- padj[81:160]
```

```{r}
View(enrich_df[enrich_df$FDR_global_pv < 0.05 |
                   enrich_df$FDR_global_py < 0.05,])
```

There are 30 significantly enriched in P. vivax and 13 significantly enriched in P. yoelii. The 11 motifs CCA, CAU, UAA, UAC, UAG, UCA, UUU, CA, GC, UA, UC, and UG are enriched in both.

```{r}
enriched_motifs_pv <- enrich_df$Motif[enrich_df$FDR_global_pv < 0.05]
enriched_motifs_py <- enrich_df$Motif[enrich_df$FDR_global_py < 0.05]
```

## Frequency of motifs in 5' UTRs and CDS

Do we get the same statistical enrichment that we see in 3' UTRs?

Extract 5' UTR sequences

```{r}
gff_pv_5utr <- gff_pv[gff_pv$type == "five_prime_UTR"]

pred_pv_5utr <- fix_widths(pred_pv_5utr, fasta_pv)

seq_pv_5utr_1 <- BSgenome::getSeq(fasta_pv, gff_pv_5utr)
names(seq_pv_5utr_1) <- sub("\\..*", "", sub("^utr_", "", gff_pv_5utr$ID))
seq_pv_5utr_2 <- BSgenome::getSeq(fasta_pv, pred_pv_5utr)
names(seq_pv_5utr_2) <- pred_pv_5utr$gene_id

seq_pv_5utr <- c(seq_pv_5utr_1, seq_pv_5utr_2)
rm(seq_pv_5utr_1, seq_pv_5utr_2)

pred_py_5utr <- fix_widths(pred_py_5utr, fasta_py)

seq_py_5utr <- BSgenome::getSeq(fasta_py, pred_py_5utr)
names(seq_py_5utr) <- pred_py_5utr$gene_id

seq_pv_5utr_split <- unlist(strsplit(seq_pv_5utr, "N"))
```

Run enrichment tests in 5' UTR and build results table

```{r}
all_tests_pv <- sapply(
    all_motifs,
    test_motif_enrichment,
    seqs = seq_pv_5utr_split
)

all_tests_py <- sapply(
    all_motifs,
    test_motif_enrichment,
    seqs = seq_py_5utr
)
```

```{r}
enrich_df_5utr <- data.frame(Motif = all_motifs,
                        Matches_pv = all_tests_pv["matches",],
                        Rate_ratio_pv = all_tests_pv["rate_ratio",],
                        Poisson_P_pv = all_tests_pv["p_value",],
                        Matches_py = all_tests_py["matches",],
                        Rate_ratio_py = all_tests_py["rate_ratio",],
                        Poisson_P_py = all_tests_py["p_value",])
```

```{r}
padj <- p.adjust(c(enrich_df_5utr$Poisson_P_pv, enrich_df_5utr$Poisson_P_py),
                 method = "fdr")

enrich_df_5utr$FDR_global_pv <- padj[1:80]
enrich_df_5utr$FDR_global_py <- padj[81:160]
```

P. vivax still has a lot of UGUANNNUA and UGUANNUA motifs in the 5' UTR.

```{r}
enriched_motifs_pv <- enrich_df_5utr$Motif[enrich_df_5utr$FDR_global_pv < 0.05]
enriched_motifs_py <- enrich_df_5utr$Motif[enrich_df_5utr$FDR_global_py < 0.05]
```

Get the sequence for spliced CDS. Yoelii has just one gene with two transcripts, but the first is longer so we will discard the second. There are still some N's in the P. vivax CDS, so take those out for searching for motifs.

```{r}
gff_pv_cds <- gff_pv[gff_pv$type == "CDS"]
gff_py_cds <- gff_py[gff_py$type == "CDS"]

seq_pv_cds <- BSgenome::getSeq(fasta_pv, gff_pv_cds)
names(seq_pv_cds) <- gff_pv_cds$ID

seq_py_cds <- BSgenome::getSeq(fasta_py, gff_py_cds)
names(seq_py_cds) <- gff_py_cds$ID

all(grepl(".1-p1", names(seq_pv_cds), fixed = TRUE)) # TRUE
all(grepl(".1-p1", names(seq_py_cds), fixed = TRUE)) # FALSE
grep(".2-p1", names(seq_py_cds), value = TRUE, fixed = TRUE)

seq_py_cds <- seq_py_cds[grep(".1-p1", names(seq_py_cds), fixed = TRUE)]

splice_seqs <- function(seq){
    stopifnot(all(grepl(".1-p1-CDS", names(seq), fixed = TRUE)))
    temp <- strsplit(names(seq), ".1-p1-CDS", fixed = TRUE)
    gene_ids <- sapply(temp, `[`, 1)
    cds_num <- as.integer(sapply(temp, `[`, 2))
    out <- sapply(
        unique(gene_ids),
        \(x){
            rw <- which(gene_ids == x)
            theseseq <- seq[rw][cds_num[rw]]
            do.call(xscat, theseseq)
        }
    ) |> DNAStringSet()
    return(out)
}

seq_pv_cds_splice <- splice_seqs(seq_pv_cds)
seq_py_cds_splice <- splice_seqs(seq_py_cds)

test <- vcountPattern("N", seq_pv_cds_splice, fixed = TRUE)
sum(test)

seq_pv_cds_split <- unlist(strsplit(seq_pv_cds_splice, "N"))
```

Run enrichment in CDS and build results table.

```{r}
all_tests_pv <- sapply(
    all_motifs,
    test_motif_enrichment,
    seqs = seq_pv_cds_split
)

all_tests_py <- sapply(
    all_motifs,
    test_motif_enrichment,
    seqs = seq_py_cds_splice
)
```

```{r}
enrich_df_cds <- data.frame(Motif = all_motifs,
                        Matches_pv = all_tests_pv["matches",],
                        Rate_ratio_pv = all_tests_pv["rate_ratio",],
                        Poisson_P_pv = all_tests_pv["p_value",],
                        Matches_py = all_tests_py["matches",],
                        Rate_ratio_py = all_tests_py["rate_ratio",],
                        Poisson_P_py = all_tests_py["p_value",])
```

Load in 3' UTR results and build one table of everything

```{r}
enrich_df_3utr <- enrich_df
```

```{r}
enrich_df <- data.frame(
    Motif = rep(all_motifs, times = 6),
    Species = rep(rep(c("*P. vivax* P01", "*P. yoelii* 17X"), each = 80), times = 3),
    Region = rep(c("3' UTR", "5' UTR", "CDS"), each = 160),
    Matches = c(
        enrich_df_3utr$Matches_pv, enrich_df_3utr$Matches_py,
        enrich_df_5utr$Matches_pv, enrich_df_5utr$Matches_py,
        enrich_df_cds$Matches_pv, enrich_df_cds$Matches_py
    ),
    Rate_ratio = c(
        enrich_df_3utr$Rate_ratio_pv, enrich_df_3utr$Rate_ratio_py,
        enrich_df_5utr$Rate_ratio_pv, enrich_df_5utr$Rate_ratio_py,
        enrich_df_cds$Rate_ratio_pv, enrich_df_cds$Rate_ratio_py
    ),
    Poisson_P = c(
        enrich_df_3utr$Poisson_P_pv, enrich_df_3utr$Poisson_P_py,
        enrich_df_5utr$Poisson_P_pv, enrich_df_5utr$Poisson_P_py,
        enrich_df_cds$Poisson_P_pv, enrich_df_cds$Poisson_P_py
    )
)

enrich_df$Global_FDR <- p.adjust(enrich_df$Poisson_P, method = "fdr")
```

```{r}
#write.table(enrich_df,
#            "../results/Motif_enrichment_2026-10-08.txt",
#            sep = "\t", row.names = FALSE, col.names = TRUE)
```

```{r}
enrich_df <- read.table("../results/Motif_enrichment_2026-10-08.txt",
                        sep = "\t", header = TRUE)
```

Look at significant hits

```{r}
enrich_df_sig <- enrich_df[enrich_df$Global_FDR < 0.05,]
enrich_df_sig$Short <- substring(enrich_df_sig$Motif, 5,
                                 nchar(enrich_df_sig$Motif) - 2)
enrich_df_sig$Region <- factor(enrich_df_sig$Region,
                               levels = c("5' UTR", "CDS", "3' UTR"))

table(enrich_df_sig$Region, enrich_df_sig$Species)

table(enrich_df_sig$Short, enrich_df_sig$Species)
```

Overall enrichment of UGUANNNUA

| Region           | Rate ratio | P-value |
|------------------|------------|---------|
| P. vivax 5' UTR  | 1.28       | 1.3e-32 |
| P. vivax CDS     | 0.98       | 0.81    |
| P. vivax 3' UTR  | 1.26       | 9.6e-26 |
| P. yoelii 5' UTR | 0.96       | 1.0     |
| P. yoelii CDS    | 0.65       | 1.0     |
| P. yoelii 3' UTR | 0.94       | 1.0     |

```{r}
temp <- test_motif_enrichment("UGUANNNUA", seq_pv_3utr_split)
temp

temp <- test_motif_enrichment("UGUANNNUA", seq_pv_5utr_split)
temp

temp <- test_motif_enrichment("UGUANNNUA", seq_pv_cds_split)
temp

temp <- test_motif_enrichment("UGUANNNUA", seq_py_3utr)
temp

temp <- test_motif_enrichment("UGUANNNUA", seq_py_5utr)
temp

temp <- test_motif_enrichment("UGUANNNUA", seq_py_cds_splice)
temp
```

## Visualize enrichment test

This is a component of Supplementary Figure 5.

```{r}
#| fig.width: 8.2
#| fig.height: 5.5
p <- enrich_df %>%
    mutate(Region = factor(Region, levels = c("5' UTR", "CDS", "3' UTR"))) %>%
    ggplot(aes(x = Matches, y = Rate_ratio)) +
    geom_point(aes(color = Global_FDR < 0.05)) +
    geom_text_repel(data = enrich_df_sig, mapping = aes(label = Short),
                    color = "red") +
    scale_color_manual(values = c("black", "red")) +
    xlab("N. motifs found") +
    ylab("Rate ratio") +
    facet_grid(cols = vars(Region), rows = vars(Species),
               space = "free_x", scale = "free") +
    theme(strip.text.y = element_markdown(),
          text = element_text(family = "Arial"))
print(p)
```

```{r}
pdf("../results/figure_components/enrichment_scatter_2026-10-08.pdf",
    width = 8.2, height = 5.5)
print(p)
dev.off()
```

## Identify most highly enriched motifs, for downstream analysis

```{r}
sig_3utr_both <- enrich_df %>%
    filter(Region == "3' UTR") %>%
    group_by(Motif) %>%
    summarize(Sig_both = all(Global_FDR < 0.05)) %>%
    filter(Sig_both) %>%
    pull(Motif)

sig_3utr_both
```

Ten motifs were significantly enriched in the 3' UTRs of both P. vivax and P. yoelii

- UGUA**CA**UA

- UGUA**GC**UA

- UGUA**UA**UA

- UGUA**UG**UA

- UGUA**CCA**UA

- UGUA**CUA**UA

- UGUA**UAA**UA

- UGUA**UAC**UA

- UGUA**UCA**UA

- UGUA**UUU**UA

```{r}
sig_utr_all <- enrich_df %>%
    filter(Region %in% c("3' UTR", "5' UTR")) %>%
    group_by(Motif) %>%
    summarize(Sig_both = all(Global_FDR < 0.05)) %>%
    filter(Sig_both) %>%
    pull(Motif)

sig_utr_all
```

Reducing it to those that were significant in both the 3' and 5' UTRs in both species, we just have the four 8-nucleotide motifs.

## Assign motifs to genes

If we just search the 3' UTR, 5' UTR, or CDS sequence directly we will have to back-calculate genomic coordinates if we want them.

```{r}
matches_by_motif_pv <- sapply(
    all_motifs,
    \(x){
        pat <- gsub("U", "T", x)
        vm <- vmatchPattern(pat, subject = seq_pv_3utr)
        df1 <- data.frame(Gene = rep(names(vm), times = lengths(vm)),
                         Position = unlist(start(vm)))
        df1$Motif <- x
        df1$Region <- "3' UTR"
        
        vm <- vmatchPattern(pat, subject = seq_pv_5utr)
        df2 <- data.frame(Gene = rep(names(vm), times = lengths(vm)),
                         Position = unlist(start(vm)))
        df2$Motif <- x
        df2$Region <- "5' UTR"
        
        vm <- vmatchPattern(pat, subject = seq_pv_cds_splice)
        df3 <- data.frame(Gene = rep(names(vm), times = lengths(vm)),
                         Position = unlist(start(vm)))
        df3$Motif <- x
        df3$Region <- "CDS"
        
        return(rbind(df1, df2, df3))
    },
    simplify = FALSE
)

matches_pv <- do.call(rbind, matches_by_motif_pv)
rm(matches_by_motif_pv)

matches_by_motif_py <- sapply(
    all_motifs,
    \(x){
        pat <- gsub("U", "T", x)
        vm <- vmatchPattern(pat, subject = seq_py_3utr)
        df1 <- data.frame(Gene = rep(names(vm), times = lengths(vm)),
                         Position = unlist(start(vm)))
        df1$Motif <- rep(x, nrow(df1))
        df1$Region <- rep("3' UTR", nrow(df1))
        
        vm <- vmatchPattern(pat, subject = seq_py_5utr)
        df2 <- data.frame(Gene = rep(names(vm), times = lengths(vm)),
                         Position = unlist(start(vm)))
        df2$Motif <- rep(x, nrow(df2))
        df2$Region <- rep("5' UTR", nrow(df2))
        
        vm <- vmatchPattern(pat, subject = seq_py_cds_splice)
        df3 <- data.frame(Gene = rep(names(vm), times = lengths(vm)),
                         Position = unlist(start(vm)))
        df3$Motif <- rep(x, nrow(df3))
        df3$Region <- rep("CDS", nrow(df3))
        
        return(rbind(df1, df2, df3))
    },
    simplify = FALSE
)

matches_py <- do.call(rbind, matches_by_motif_py)
rm(matches_by_motif_py)
```

List out genes by motif.

```{r}
genes_by_motif_pv_txpt <- split(matches_pv$Gene, matches_pv$Motif)
genes_by_motif_py_txpt <- split(matches_py$Gene, matches_py$Motif)

temp <- matches_pv$Region == "3' UTR"
genes_by_motif_pv_3utr <- split(matches_pv$Gene[temp],
                                matches_pv$Motif[temp])
temp <- matches_py$Region == "3' UTR"
genes_by_motif_py_3utr <- split(matches_py$Gene[temp],
                                matches_py$Motif[temp])
```

```{r}
# save(matches_pv, matches_py,
#      genes_by_motif_pv_txpt, genes_by_motif_py_txpt,
#      genes_by_motif_pv_3utr, genes_by_motif_py_3utr,
#      file = "../results/R_objects/matches_2026-10-08.RData")
```

## Motif enrichment in DGE

```{r}
load("../results/R_objects/matches_2026-10-08.RData")

temp <- matches_pv$Region == "5' UTR"
genes_by_motif_pv_5utr <- split(matches_pv$Gene[temp],
                                matches_pv$Motif[temp])
temp <- matches_py$Region == "5' UTR"
genes_by_motif_py_5utr <- split(matches_py$Gene[temp],
                                matches_py$Motif[temp])
```

```{r}
res_pv <- read.table("../results/vivax_dge_results.txt",
                     sep = "\t", header = TRUE)
```

Set up data frame to contain output for all motifs, in the two UTR regions.

```{r}
dge_enrich_df <- data.frame(
    Motif = rep(all_motifs, times = 2),
    Region = rep(c("3' UTR", "5' UTR"), each = length(all_motifs))
)

dge_enrich_df$Mean_logFC_with <- NA_real_
dge_enrich_df$Mean_logFC_without <- NA_real_
dge_enrich_df$Pval_Ttest_twosided <- NA_real_
dge_enrich_df$N_genes <- NA_integer_
```

Loop through motifs and fill in results

```{r}
for(motif in all_motifs){
    for(reg in c("3' UTR", "5' UTR")){
        if(reg == "3' UTR"){
            gbm <- genes_by_motif_pv_3utr
        } else {
            gbm <- genes_by_motif_pv_5utr
        }
        temprows <- which(res_pv$Gene %in% gbm[[motif]])
        logFC_a <- res_pv$logFC[temprows]
        logFC_b <- res_pv$logFC[-temprows]
        i <- which(dge_enrich_df$Motif == motif & dge_enrich_df$Region == reg)
        dge_enrich_df$Mean_logFC_with[i] <- mean(logFC_a)
        dge_enrich_df$Mean_logFC_without[i] <- mean(logFC_b)
        dge_enrich_df$N_genes[i] <- length(temprows)
        
        if(length(temprows) > 2){
            myT <- t.test(logFC_a, logFC_b, alternative = "two.sided")
            dge_enrich_df$Pval_Ttest_twosided[i] <- myT$p.value
        }
    }
}

dge_enrich_df$Global_FDR <- p.adjust(dge_enrich_df$Pval_Ttest_twosided,
                                     method = "fdr")
```

All that are significant show positive enrichment with hypnozoites. UGUA**CAU**UA in the 3' UTR is the most significant, but not the only. There is also UGUA**AAU**UA, UGUA**ACU**UA, UGUA**CUU**UA (3' UTR), UGUA**AAC**UA, UGUA**ACC**UA, and UGUA**CUA**UA (5' UTR) for the nine-nucleotide motifs, and UGUA**AG**UA in 5' UTR. This set of nine-nucleotide motifs is not exactly the same set that we saw enriched in Pv and Py 3' UTRs but what is similar is that they all lack G's in the variable region.

```{r}
dge_enrich_df %>%
    filter(Global_FDR < 0.05)
```

```{r}
#write.table(dge_enrich_df,
#            file = "../results/DGE_enrichment_2026-10-09.txt",
#            sep = "\t", row.names = FALSE, col.names = TRUE)
```

## Visualize sequence enrichment and DGE together

Compare sequence enrichment with differential gene expression

```{r}
combined_df <- enrich_df %>%
    filter(Species == "*P. vivax* P01" & Region != "CDS") %>%
    left_join(dge_enrich_df, by = c("Motif", "Region")) %>%
    mutate(Short = substring(Motif, 5, nchar(Motif) - 2))
```

```{r}
ggplot(combined_df, aes(x = Rate_ratio, y = Mean_logFC_with)) +
    geom_label(aes(label = Short)) +
    facet_wrap(vars(Region))
```

```{r}
#| fig.width: 7
#| fig.height: 10
combined_df %>%
    filter(grepl("^[UAC]{3}$", Short)) %>%
ggplot(aes(x = Rate_ratio, y = Mean_logFC_with)) +
    geom_hline(yintercept = mean(res_pv$logFC), color = "blue") +
    geom_vline(xintercept = 1, color = "blue") +
    geom_label(aes(label = Short,
                   fill = Global_FDR.x < 0.05 & Global_FDR.y < 0.05)) +
    facet_wrap(vars(Region), ncol = 1) +
    labs(fill = "Sig. both tests") +
    scale_fill_manual(values = c("white", "lightgreen"))
```

Build a grid to show concordance across motifs

```{r}
combined_df$FDR_enriched_Pv <- "≥0.2"
combined_df$FDR_enriched_Pv[combined_df$Global_FDR.x < 0.2] <- "<0.2"
combined_df$FDR_enriched_Pv[combined_df$Global_FDR.x < 0.05] <- "<0.05"

combined_df$FDR_DGE_Pv <- "≥0.2"
combined_df$FDR_DGE_Pv[combined_df$Global_FDR.y < 0.2] <- "<0.2"
combined_df$FDR_DGE_Pv[combined_df$Global_FDR.y < 0.05] <- "<0.05"
combined_df$FDR_DGE_Pv[is.na(combined_df$Global_FDR.y)] <- NA

yoelii_df <- filter(enrich_df, Species == "*P. yoelii* 17X" &
                        Region %in% c("3' UTR", "5' UTR"))
yoelii_df$FDR_enriched_Py <- "≥0.2"
yoelii_df$FDR_enriched_Py[yoelii_df$Global_FDR < 0.2] <- "<0.2"
yoelii_df$FDR_enriched_Py[yoelii_df$Global_FDR < 0.05] <- "<0.05"

grid_prep <- combined_df %>%
    dplyr::select(Motif, Region, FDR_enriched_Pv, FDR_DGE_Pv) %>%
    left_join(select(yoelii_df, Motif, Region, FDR_enriched_Py),
              by = c("Motif", "Region")) %>%
    tidyr::pivot_longer(cols = starts_with("FDR_"), names_prefix = "FDR_",
                        names_to = "Test", values_to = "FDR") %>%
    mutate(Test1 = paste(Region, Test),
           Short = substring(Motif, 5, nchar(Motif) - 2))

grid_prep$Motif_category <- "9nuc with G"
grid_prep$Motif_category[nchar(grid_prep$Motif) == 8] <- "8nuc"
grid_prep$Motif_category[grep("^UGUA[UCA]{3}UA$", grid_prep$Motif)] <- "9nuc HHH"
```

```{r}
#| fig.width: 6
#| fig.height: 12.5
p <- grid_prep %>%
    filter(!is.na(FDR)) %>%
ggplot(aes(x = Test, y = Motif)) +
    geom_tile(aes(fill = FDR), width = 1, height = 1, color = "black") +
    geom_text(aes(label = Short), data = filter(grid_prep, FDR == "<0.05")) +
    facet_grid(rows = vars(Motif_category), cols = vars(Region),
               scales = "free_y", space = "free_y") +
    scale_fill_manual(values = c("≥0.2" = "white", "<0.2" = "lightblue",
                                 "<0.05" = "lightgreen")) +
    theme(axis.text.x = element_text(angle = 45, hjust = 1))
print(p)
```

```{r}
pdf("../results/figure_components/motif_significance_grid_2026-10-09.pdf",
    width = 6, height = 12.5)
print(p)
dev.off()
```

## Volcano plot

Redraw the volcano plot, showing if there is UGUACAUUA in the 3' UTR of the transcript. This is a component of Figure 5.

```{r}
#| fig.width: 4
#| fig.height: 3.5
res_pv$UGUACAUUA_3UTR <- grepl("UGUACAUUA", res_pv$Motif_3UTR)

pcutoff <- max(res_pv$PValue[res_pv$FDR < 0.05])

p <- res_pv %>%
    arrange(UGUACAUUA_3UTR) %>%
    ggplot(aes(x = -log10(PValue), y = logFC)) +
    geom_vline(xintercept = -log10(pcutoff), lty = 2) +
    geom_hline(yintercept = c(-1, 1), lty = 2) +
    geom_point(aes(color = UGUACAUUA_3UTR)) +
    annotate("text", x = 7.5, y = c(2, -2),
             label = c("Hypnozoite-enriched", "Schizont-enriched")) +
    scale_color_manual(values = c("black", "red"), guide = "none") +
    annotate("text", x = 5.2, y = 5.2, label = "Red =\nUGUACAUUA\nin 3' UTR",
             color = "red") +
    theme_classic(base_size = 10, base_family = "Arial") +
    theme(legend.position = "inside",
          legend.position.inside = c(0.8, 0.87),
          legend.background = element_rect(color = "black", linetype = "solid"))

print(p)
```

```{r}
pdf("../results/figure_components/volcano_2026-06-10.pdf",
    height = 3.35, width = 3.82)
print(p)
dev.off()
```

```{r}
svg("../results/figure_components/volcano_2026-06-10.svg",
    height = 3.35, width = 3.82)
print(p)
dev.off()
```

Ranking of genes by logFC, labeling the genes with UGUACAUUA. This is a component of Supplemental Figure 5.

```{r}
res_pv$Rank <- rank(res_pv$logFC)

midrank <- nrow(res_pv) / 2

hghclr <- "red"

genes1 <-
    res_pv %>%
    filter(Rank > midrank & logFC > 2.5 & UGUACAUUA_3UTR) %>%
    arrange(desc(Rank)) %>%
    pull(Gene)

genes2 <-
    res_pv %>%
    filter(Rank > 2950 & logFC < 2.5 & UGUACAUUA_3UTR) %>%
    arrange(desc(Rank)) %>%
    pull(Gene)

genes3 <-
    res_pv %>%
    filter(Rank > midrank & logFC > 0 & UGUACAUUA_3UTR &
               !Gene %in% c(genes1, genes2)) %>%
    arrange(desc(Rank)) %>%
    pull(Gene)

genes4 <-
    res_pv %>%
    filter(logFC < 0 & logFC > -0.3 & UGUACAUUA_3UTR) %>%
    arrange(desc(Rank)) %>%
    pull(Gene)

genes5 <-
    res_pv %>%
    filter(logFC < -0.3 & logFC > -0.4 & UGUACAUUA_3UTR) %>%
    arrange(desc(Rank)) %>%
    pull(Gene)

genes6 <- res_pv$Gene[res_pv$Rank == 1675]

genes7 <-
    res_pv %>%
    filter(Rank < midrank & Rank > 1200 & UGUACAUUA_3UTR) %>%
    arrange(desc(Rank)) %>%
    pull(Gene)

genes8 <-
    res_pv %>%
    filter(Rank < 1200 & Rank > 800 & UGUACAUUA_3UTR) %>%
    arrange(desc(Rank)) %>%
    pull(Gene)

genes9 <-
    res_pv %>%
    filter(Rank < 800 & UGUACAUUA_3UTR) %>%
    arrange(desc(Rank)) %>%
    pull(Gene)
```

```{r}
#| fig.width: 6
#| fig.height: 6
p <- res_pv %>%
    arrange(UGUACAUUA_3UTR) %>%
ggplot(aes(x = Rank, y = logFC)) +
    geom_hline(yintercept = 0) +
    geom_vline(xintercept = midrank) +
    
    annotate("segment", x = 3200, y = 6.5, xend = 3316, yend = 3.82,
             color = hghclr, lwd = 1) +
    annotate("label", x = 2900, y = 7.5, #fill = hghclr,
             label = paste(genes1, collapse = "\n")) +
    
    annotate("segment", x = 2800, y = 3.3, xend = 3201, yend = 1.83,
             color = hghclr, lwd = 1) +
    annotate("label", x = 2800, y = 4.5, #fill = hghclr,
             label = paste(genes2, collapse = "\n")) +
    
    annotate("segment", x = 2300, y = 1.5, xend = 2906, yend = 0.97,
             color = hghclr, lwd = 1) +
    annotate("label", x = 2000, y = 4, #fill = hghclr,
             label = paste(genes3, collapse = "\n")) +
    
    annotate("segment", x = 2750, y = -0.7, xend = 2195, yend = -0.11,
             color = hghclr, lwd = 1) +
    annotate("label", x = 3100, y = -2, #fill = hghclr,
             label = paste(genes4, collapse = "\n")) +
    
    annotate("segment", x = 2100, y = -1.3, xend = 1991, yend = -0.34,
             color = hghclr, lwd = 1) +
    annotate("label", x = 2250, y = -2.5, #fill = hghclr,
             label = paste(genes5, collapse = "\n")) +
    
    annotate("segment", x = 1800, y = -5, xend = 1675, yend = -0.72,
             color = hghclr, lwd = 1) +
    annotate("label", x = 2100, y = -5, #fill = hghclr,
             label = genes6) +
    
    annotate("segment", x = 1400, y = -3.5, xend = 1414, yend = -1.08,
             color = hghclr, lwd = 1) +
    annotate("label", x = 1250, y = -4, #fill = hghclr,
             label = paste(genes7, collapse = "\n")) +
    
    annotate("segment", x = 650, y = -1.5, xend = 922, yend = -1.965,
             color = hghclr, lwd = 1) +
    annotate("label", x = 300, y = -1.5, #fill = hghclr,
             label = paste(genes8, collapse = "\n")) +
    
    annotate("segment", x = 900, y = -6.6, xend = 421, yend = -5.28,
             color = hghclr, lwd = 1) +
    annotate("label", x = 900, y = -7, #fill = hghclr,
             label = genes9) +
    
    annotate("text", x = 800, y = 5, label = "Red = UGUACAUUA\nin 3' UTR",
             color = hghclr) +
    annotate("text", x = 2300, y = 9.5, label = "Hypnozoite-enriched") +
    annotate("text", x = 700, y = -12, label = "Schizont-enriched") +
    
    geom_point(aes(color = UGUACAUUA_3UTR), size = 2) +
    scale_color_manual(values = c("black", hghclr), guide = "none") +
    theme_classic(base_size = 10, base_family = "Arial")

print(p)
```

```{r}
pdf("../results/figure_components/dge_ranking_UGUACAUUA_2026-06-11a.pdf",
    width = 6, height = 6)
print(p)
dev.off()
```

## Venn diagram

Just look within expressed orthologs.

```{r}
expr_ortho_vvx <- ortho_df$Vivax_gene[ortho_df$Expressed_both]

expr_ortho_yli <- ortho_df$Yoelii_gene[ortho_df$Expressed_both]
```

Get names of genes with motifs of interest.

```{r}
motifs_of_interest <- paste0(
    "UGUA",
    c("AAA", "AAU", "AUA", "AUU", "UAA", "UAU", "UUA", "UUU", "CAU"),
    "UA"
)

motif_genes_pv_txpt <- genes_by_motif_pv_txpt[motifs_of_interest] |> unlist() |> unique()

motif_genes_py_txpt <- genes_by_motif_py_txpt[motifs_of_interest] |> unlist() |> unique()
```

Out of these 3511 expressed orthologs, 496 have motifs in both species, 153 have motifs in vivax only, 1941 have motifs in yoelii only, and 921 have motifs in neither. There is significantly more overlap than expected under the null hypothesis.

```{r}
sum(expr_ortho_vvx %in% motif_genes_pv_txpt &
        expr_ortho_yli %in% motif_genes_py_txpt)

sum(expr_ortho_vvx %in% motif_genes_pv_txpt &
        !expr_ortho_yli %in% motif_genes_py_txpt)

sum(!expr_ortho_vvx %in% motif_genes_pv_txpt &
        expr_ortho_yli %in% motif_genes_py_txpt)

sum(!expr_ortho_vvx %in% motif_genes_pv_txpt &
        !expr_ortho_yli %in% motif_genes_py_txpt)

chisq.test(matrix(c(496, 153, 1941, 921), nrow = 2, ncol = 2))
fisher.test(matrix(c(496, 153, 1941, 921), nrow = 2, ncol = 2))
```

Draw the Venn diagram that is part of Figure 5.

```{r}
#| fig.height: 7
#| fig.width: 7
scl <- 0.65
r_vvx <- sqrt(496+153) * scl
r_yli <- sqrt(496+1941) * scl
r_all <- sqrt(3511) * scl

#pdf("../results/figure_components/venn_2026-06-11.pdf",
#    width = 7, height = 7)
#svg("../results/figure_components/venn_2026-06-11.svg",
#    width = 7, height = 7)
plot(c(0, 0, 101, 101), c(0, 101, 0, 101), col = "white", col.axis = "white",
     col.lab = "white", col.main = "white",
     fg = "white")
draw.circle(50, 29, radius = r_vvx, lwd = 3, col = "#00000025")
draw.circle(50, 55, radius = r_yli, lwd = 3, col = "#00000025")
draw.circle(50, 50, radius = r_all, lwd = 3)
text(50, c(14.5, 30, 70), c("153", "496", "1941"), cex = 2, family = "Arial")
text(50, 3, expression(paste(italic("P. vivax"), " P01")), cex = 2, family = "Arial")
text(50, 97, expression(paste(italic("P. yoelii"), " 17X")), cex = 2, family = "Arial")
text(72.1, 21.5, "921", cex = 2, family = "Arial")
#dev.off()
```

## Overrepresentation analysis

```{r}
import_go <- function(pattern, dir = "../json_veupathdb"){
    fls <- list.files(path = dir, pattern = pattern)
    grp <- sub(".*_", "", sub("\\.json", "", fls))
    lst <- lapply(fls,
                  \(x) sapply(read_json(paste0(dir, "/", x)), unlist))
    for(i in seq_along(fls)){
        names(lst[[i]]) <- paste(grp[i], names(lst[[i]]), sep = ":")
    }
    return(do.call(c, lst))
}
```

```{r}
Pv_go_anno <- import_go("PvivaxP01_GO_annotated_.*\\.json")
Py_go_anno <- import_go("Pyoelii17X_GO_annotated_.*\\.json")
```

```{r}
sub_fora <- function(df, n = 10){
    df %>%
        arrange(pval) %>%
        head(n) %>%
        mutate(Ontology = sub("\\:.*", "", pathway)) %>%
        mutate(Pathway = sub(".*\\:", "", pathway)) %>%
        select(Ontology, Pathway, pval, overlap, foldEnrichment)
}
fora_gg_prep <- function(fora_pv, fora_py, n = 10){
    df1 <- sub_fora(fora_pv, n = n)
    df1$Species <- "*P. vivax* P01"
    df2 <- sub_fora(fora_py, n = n) 
    df2$Species <- "*P. yoelii* P17X"
    return(rbind(df1, df2))
}

lollipop <- function(df){
    dummy <- as.character(seq_len(nrow(df)))
    df$Dummy <- factor(dummy,
                       levels = dummy[order(df$Ontology, df$foldEnrichment)])
    ggplot(df, aes(y = Dummy)) +
        geom_segment(aes(x = 0, xend = foldEnrichment,
                         yend = Dummy)) +
        geom_point(aes(x = foldEnrichment, size = overlap,
                       fill = -log10(pval)),
                   shape = 21) +
        geom_text(aes(x = 0, label = Pathway, color = Ontology),
                  nudge_y = 0.3, hjust = 0) +
        xlab("Fold enrichment") + ylab("Pathway") +
        labs(size = "Overlap") +
        scale_fill_viridis_c() +
        facet_wrap(vars(Species), scales = "free") +
        theme_classic() +
        theme(axis.text.y = element_blank(),
              plot.margin = margin(l = 0.9, unit = "cm"))
}
```

### Either motif anywhere in the transcript

Using the gene sets from the Venn diagram.

```{r}
fora_pv_only_cau_www_txpt <- fora(
    Pv_go_anno,
    expr_ortho_vvx[expr_ortho_vvx %in% motif_genes_pv_txpt &
        !expr_ortho_yli %in% motif_genes_py_txpt],
    expr_ortho_vvx)

fora_pv_both_cau_www_txpt <- fora(
    Pv_go_anno,
    expr_ortho_vvx[expr_ortho_vvx %in% motif_genes_pv_txpt &
        expr_ortho_yli %in% motif_genes_py_txpt],
    expr_ortho_vvx)

fora_py_only_cau_www_txpt <- fora(
    Py_go_anno,
    expr_ortho_yli[!expr_ortho_vvx %in% motif_genes_pv_txpt &
        expr_ortho_yli %in% motif_genes_py_txpt &
            !duplicated(expr_ortho_yli)],
    expr_ortho_yli[!duplicated(expr_ortho_yli)])

fora_py_both_cau_www_txpt <- fora(
    Py_go_anno,
    expr_ortho_yli[expr_ortho_vvx %in% motif_genes_pv_txpt &
        expr_ortho_yli %in% motif_genes_py_txpt &
            !duplicated(expr_ortho_yli)],
    expr_ortho_yli[!duplicated(expr_ortho_yli)])

#save(fora_pv_both_cau_www_txpt,
#     fora_pv_only_cau_www_txpt,
#     fora_py_both_cau_www_txpt,
#     fora_py_only_cau_www_txpt,
#     file = "../results/R_objects/fora_2026-06-10.RData")
```

```{r}
load("../results/R_objects/fora_2026-06-10.RData")
```

Prep for ggplot

```{r}
temp1 <- sub_fora(fora_pv_only_cau_www_txpt)
temp1$Species <- "*P. vivax* P01"
temp1$Orthology <- "Exclusive"

temp2 <- sub_fora(fora_pv_both_cau_www_txpt)
temp2$Species <- "*P. vivax* P01"
temp2$Orthology <- "Overlapping"

temp3 <- sub_fora(fora_py_only_cau_www_txpt)
temp3$Species <- "*P. yoelii* P17X"
temp3$Orthology <- "Exclusive"

temp4 <- sub_fora(fora_py_both_cau_www_txpt)
temp4$Species <- "*P. yoelii* P17X"
temp4$Orthology <- "Overlapping"

plot_df <- rbind(temp1, temp2, temp3)#, temp4)

dummy <- as.character(seq_len(nrow(plot_df)))
plot_df$Dummy <- factor(dummy,
                       levels = dummy[order(plot_df$Ontology, plot_df$foldEnrichment)])
```

Lollipop plot that is a component of Figure 5.

```{r}
#| fig-width: 5
#| fig-height: 6
p <- ggplot(plot_df, aes(y = Dummy)) +
    geom_segment(aes(x = 0, xend = foldEnrichment,
                     yend = Dummy)) +
    geom_point(aes(x = foldEnrichment, size = overlap,
                   fill = -log10(pval)),
               shape = 21) +
    geom_text(aes(x = 0, label = Pathway, color = Ontology),
              nudge_y = 0.3, hjust = 0, size = 8 / .pt) +
    xlab("Fold enrichment") + ylab("Pathway") +
    labs(size = "Overlap") +
    scale_fill_viridis_c() +
    facet_wrap(vars(paste(Species, Orthology)), scales = "free_y", ncol = 1) +
    theme_classic(base_size = 10, base_family = "Arial") +
    theme(axis.text.y = element_blank(), axis.ticks.y = element_blank(),
          #plot.margin = margin(l = 0.9, unit = "cm"),
          strip.text.x = element_markdown())
print(p)
```

```{r}
pdf("../results/figure_components/lollipops_2026-06-11.pdf", width = 5, height = 6)
print(p)
dev.off()
```

```{r}
svg("../results/figure_components/lollipops_2026-06-11.svg", width = 5, height = 6)
print(p)
dev.off()
```

## Barchart of gene counts

```{r}
temp1 <- matches_pv %>%
    filter(Gene %in% expr_ortho_vvx) %>%
    filter(Motif %in% motifs_of_interest) %>%
    group_by(Region, Gene) %>%
    summarize(Motifs_per_gene = n()) %>%
    summarize(One_motif = sum(Motifs_per_gene == 1),
              Multi_motif = sum(Motifs_per_gene > 1)) %>%
    tidyr::pivot_longer(cols = c("One_motif", "Multi_motif"),
                        names_to = "Number_motifs",
                        values_to = "Number_of_transcripts")
temp1$Species <- "*P. vivax* P01"

temp2 <- matches_py %>%
    filter(Gene %in% expr_ortho_yli) %>%
    filter(Motif %in% motifs_of_interest) %>%
    group_by(Region, Gene) %>%
    summarize(Motifs_per_gene = n()) %>%
    summarize(One_motif = sum(Motifs_per_gene == 1),
              Multi_motif = sum(Motifs_per_gene > 1)) %>%
    tidyr::pivot_longer(cols = c("One_motif", "Multi_motif"),
                        names_to = "Number_motifs",
                        values_to = "Number_of_transcripts")
temp2$Species <- "*P. yoelii* P17X"

plot_df <- rbind(temp1, temp2)
plot_df$Number_motifs[plot_df$Number_motifs == "One_motif"] <- "1"
plot_df$Number_motifs[plot_df$Number_motifs == "Multi_motif"] <- ">1"

anno_df <- plot_df %>%
    group_by(Region, Species) %>%
    summarize(Total = sum(Number_of_transcripts))
# Dummy variable for x-axis placement
anno_df$Number_of_transcripts <- rep(c(100, 500), times = 3)
```

This barchart is a component of Figure 5.

```{r}
#| fig.width: 3
#| fig.height: 2
p <- ggplot(plot_df, aes(x = Region, y = Number_of_transcripts)) +
    geom_col(aes(fill = Number_motifs)) +
    geom_text(data = anno_df, mapping = aes(label = Total)) +
    facet_wrap(vars(Species), scales = "free_x") +
    scale_fill_manual(values = RColorBrewer::brewer.pal(2, "Paired")[2:1]) +
    coord_flip() +
    labs(fill = "Motifs per gene", y = "Number of genes") +
    theme_classic(base_family = "Arial", base_size = 10) +
    theme(strip.text.x = element_markdown(), legend.position = "bottom")
print(p)
```

```{r}
pdf("../results/figure_components/barplot_gene_region_2026-06-10.pdf",
    height = 2, width = 3)
print(p)
dev.off()
```

```{r}
svg("../results/figure_components/barplot_gene_region_2026-06-10.svg",
    height = 2, width = 3)
print(p)
dev.off()
```

## Tables to export

```{r}
ortho_df$Motif_vivax <- ortho_df$Vivax_gene %in% motif_genes_pv_txpt
ortho_df$Motif_yoelii <- ortho_df$Yoelii_gene %in% motif_genes_py_txpt
```

```{r}
write.csv(ortho_df, file = "../results/orthologs_2026-06-11.csv")
```

```{r}
write.csv(matches_pv, file = "../results/matches_pv_2026-06-11.csv",
          row.names = FALSE)
write.csv(matches_py, file = "../results/matches_py_2026-06-11.csv",
          row.names = FALSE)
```

Overrepresentation analysis

```{r}
fora_pv_only_cau_www_txpt$Species <- "P. vivax P01"
fora_pv_only_cau_www_txpt$Motifs_found_in <- "P. vivax only"

fora_pv_both_cau_www_txpt$Species <- "P. vivax P01"
fora_pv_both_cau_www_txpt$Motifs_found_in <- "Both species"

fora_py_only_cau_www_txpt$Species <- "P. yoelii 17X"
fora_py_only_cau_www_txpt$Motifs_found_in <- "P. yoelii only"

fora_py_both_cau_www_txpt$Species <- "P. yoelii 17X"
fora_py_both_cau_www_txpt$Motifs_found_in <- "Both species"

rbind(fora_pv_only_cau_www_txpt,
      fora_pv_both_cau_www_txpt,
      fora_py_only_cau_www_txpt,
      fora_py_both_cau_www_txpt) %>%
    mutate(overlapGenes = sapply(overlapGenes, paste, collapse = ";")) %>%
    write.table(file = "../results/fora_2026-06-11.txt", sep = "\t",
                row.names = FALSE, col.names = TRUE)
```
